---
title: 'DATA 624: Homework 2'
subtitle: 'Time Series Decomposition'
author: "Andrew Bowen"
date: "2024-01-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries, echo=FALSE, message=FALSE}
library(tidyverse)
library(tsibble)
library(fpp3)
```



## Exercise 3.1
First, le'ts take a look at our `global_economy` dataset:
```{r}
head(global_economy)
```


First, we'll need to calculate the GDP per capita by time period
```{r}
global_economy <- global_economy %>% mutate(gdp_per_capita = GDP / Population)

# Plot GDP per capita per country over time
autoplot(global_economy, gdp_per_capita)
```
Now let's find the countrieswith the highest GDP per capita. Looking at our sorted result, it looks like micro-nations like Monaco, Liechenstein, and Luxembourg top the list.
```{r}
# Find GDP per Capita by Country, then sort on result
highest_gdps <- global_economy %>%
              group_by(Country) %>%
              arrange(desc(gdp_per_capita))
highest_gdps
```

Not we can plot our results, let's look at Monaco's GDP per capita over time. We see a strong upward trend since the start of our dataset (1970)

```{r}
monaco <- global_economy %>%
                filter(Country == "Monaco")

# Plot results for Monaco
autoplot(monaco) + labs(x="Year", y="GDP per Capita (USD)", title="Monaco GDP per Capita over Time")
```

## Exercise 3.2


## Exercise 3.3




## Exercise 3.8


