---
title: 'DATA 624: Homework 5'
author: "Andrew Bowen"
date: "2024-02-25"
output: html_document
---

```{r setup## Exerciseinclude=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE## Exerciseecho=FALSE## Exercisewarning=FALSE}
library(tsibble)
library(tidyverse)
library(feasts)
library(fpp3)
library(fabletools)
library(tsibbledata)
```


## Exercise 8.1
First, let's construct our time series for pigs slaughtere din Victoria and plot
```{r}
victoria_pigs <- aus_livestock %>% filter(Animal == "Pigs" & State == "Victoria")

victoria_pigs %>% autoplot(Count)
```

There doesn't appread to be strong seasonality or a general overall trend in this time series, so simple exponential smoothing makes sense to use. Now, let's construct a model via the ETS function for SES. We'll mirror the techniques used in Hyndman
```{r}
# Fit model with only an error term to reflect SES
fit <- victoria_pigs %>%
  model(ETS(Count ~ error("A") + trend("N") + season("N")))
fc <- fit |>
  forecast(h = 4)
```


Plotting our forecast over our time series gives us the following
```{r}
# Plot victorian pigs model
fc |>
  autoplot(victoria_pigs) +
  geom_line(aes(y = .fitted), col="#D55E00",
            data = augment(fit)) +
  labs(y="Number of Pigs", title="Victorian Pigs Slaughtered") +
  guides(colour = "none")
```

To find the optimized values fo $\alpha$ and $l_0$, we can use the `report` function on our `fit` mable. We see optimized values for each parameter in our SES model below:
```{r}
fit %>% report()
```


## Exercise 8.5


## Exercise 8.6


## Exercise 8.7


## Exercise 8.8


## Exercise 8.9







