---
title: 'DATA 624: Homework 4'
author: "Andrew Bowen"
date: "2024-02-17"
output: html_document
---



```{r message=FALSE, echo=FALSE}
library(tidyverse)
library(fpp3)
library(fabletools)
library(mlbench)
library(tsibble)
library(corrplot)
library(GGally)
```



## Exercise 3.1 (*K&J*)

```{r}
data(Glass)
head(Glass)
```

We can view the relationship between variables and our predictor (`Type`) via a correlation plot from the `corrplot` package in R
```{r}
# First, calculate the correlatino between variables in 
glass <- Glass %>% mutate(Type = as.numeric(Type))
corrGlass <- cor(glass)
corrplot(corrGlass)
```
We see reasonably strong correlations between our `Type` variable and the predictors for percentage of Barium, Alimunum, Magnesium, and Sodium.

We can also use a corner plot from the `GGally::ggpairs` function to visually inspect the relationships between our variables. We can also gain a sense of the distributions of each variable from this plot. I'm going to remove the correlations (`upper=NULL`; see above) to make the plot a bit less busy:
```{r warning=FALSE, message=FALSE}
# Use GGally to plot pairwise relationships between variables
ggpairs(glass, upper=NULL)
```

The predictor variables for the minerals we care about from above seem to be a bit skewed from this visual. There seem to be a few outlier points for the potassium (`K`) predictor, but this variable doesn't correlate strongly with our outcome variable.

## Exercise 3.2 (*K&J*)